---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  fig.height = 7,
  fig.width=10
)
```

# gamworkshop

<!-- badges: start -->
<!-- badges: end -->

The goal of gamworkshop is to ...

```{r setup}
library(gamworkshop)
library(pacman)

p_load(data.table,tidyverse,knitr,splines,ggfortify,mgcv,ggrepel,mgcViz)

stage_knots_in_age <- c(27/365,1,2,5,11,18,21)

theme_set(theme_bw(base_size = 16) + theme(text = element_text(face="bold")))

```

# Punchline 

GAM > GLM > Logistic Regression for understanding nonlinear relationships. 

# Introduction

Oftentimes we want to model the relationship between a characteristic and an outcome. In my work, I want to evaluate a characteristic's relationship to the outcome in a way that doesn't bias the type of relationship it naturally may have. In this tutorial, I will show how generalized additive models are a flexible yet interpretable way to model relationships in a population and have advantages over traditionally used methods. 

# Outline

* Question

    - Setup the question to be investigated.
   
* Data

    - Display the data to be used to answer the question

* Relationship between outcome and characteristic

    - Show different ways of visualizing the data as well as the relationship between the outcome and characteristic.
   
* Modeling

    - Different modeling strategies to investigate the relationship. Each strategy or method will be generally explained with examples. 
    
    + Logistic Regression

    + Regression Spline

    + Generalized Additive Model
    
# Question

Which population model is quantitatively better at associating an adverse event with a drug during child development i.e. along child developmental stages?

[Child developmental stages](https://www.researchgate.net/profile/Katrina_Williams8/publication/230612139/figure/tbl2/AS:601671696650242@1520461119208/Age-Stages-Defined-According-to-NICHD-Pediatric-Terminology.png)

    
# Data

```{r data}

data("report_dat") 
report_dat$D <- factor(report_dat$D,levels=c("Other drugs","Drug"))
report_dat$E <- factor(report_dat$E,levels=c("Other events","Event"))

report_dat <- 
  report_dat %>% 
  .[order(age)]

stages = c("term_neonatal","infancy","toddler","early_childhood","middle_childhood","early_adolescence","late_adolescence")
report_dat$nichd <- factor(report_dat$nichd,levels=stages)
report_dat$nichd_int <- report_dat$nichd %>% as.integer()

report_dat[1:5] %>% 
    kable("simple")

```

# Relationship between Paranoia and report characteristics

```{r relationships}

report_dat[,.N,.(D,E,year)] %>% 
  ggplot(aes(year,N)) +
  geom_point() +
  geom_line() +
  facet_grid(E~D) +
  scale_y_continuous(trans="log10",label=scales::comma) +
  ylab("Number of reports") +
  xlab("Age")

report_dat[,.N,.(nichd,D,E)] %>% 
  ggplot(aes(N,nichd)) +
  geom_bar(stat="identity") +
  geom_text_repel(aes(label=N),nudge_x=0.1) +
  facet_grid(D~E,scales="free") +
  scale_x_continuous(trans="log10",label=scales::comma) +
  xlab("Number of reports") +
  ylab("")

tab <- report_dat[,table(D,E)]
prr <- 
  (tab["Drug","Event"] / ( tab["Drug","Event"] + tab["Drug","Other events"] ) ) / 
  (tab["Other drugs","Event"] / (tab["Other drugs","Event"] + tab["Other drugs","Other events"]) )
prr

tab <- report_dat[,table(D,E)]
prr <- function(tab){
  
  (tab["Drug","Event"] / ( tab["Drug","Event"] + tab["Drug","Other events"] ) ) / 
  (tab["Other drugs","Event"] / (tab["Other drugs","Event"] + tab["Other drugs","Other events"]) )
  
}

prr(tab)

prr_df <- 
data.table(
  nichd = stages,
  PRR = sapply(stages,function(x){prr(report_dat[nichd==x,table(D,E)])})
) 

prr_df %>% 
  ggplot(aes(factor(nichd,levels=stages),PRR)) +
  geom_bar(stat="identity") +
  xlab("")

```

# Modeling

## Logistic Regression

```{r logit}

glm_coef <- function(form="E ~ D - 1",id="id"){
    formula=as.formula(form)
    mod <- glm(formula,data=report_dat,family=binomial(link="logit"))
    summ <- summary(mod)
    dt <- summ$coefficients %>% data.table()
    dt$formula <- form
    dt$term <- summ$coefficients %>% rownames() %>% str_replace("D","") %>% str_replace("nichd",'')
    dt$id=id
    dt$aic <- mod$aic
    dt
}

glm_coefs <- 
    bind_rows(
        glm_coef("E ~ D",id='model1'),
        glm_coef("E ~ nichd",id='model2'),
        glm_coef("E ~ D + nichd",id='model3'),
        glm_coef("E ~ D*nichd",id='model4')
        )

glm_coefs$size = ifelse(stringr::str_detect(glm_coefs$term,":"),2,1)

glm_coefs %>% 
  ggplot(aes(formula,Estimate,color=term,group=term)) +
  geom_point(aes(size=size)) +
  geom_line() +
  scale_size(guide="none")

glm_coefs[,.(nterms = .N),.(formula,aic)] %>% 
  ggplot(aes(formula,aic)) +
  geom_point(size=4) +
  ylab("AIC")


age_coef <- glm_coef("E ~ age",id='model2')[term=="age",Estimate]
age_coef2 <- glm_coef("E ~ age + D",id='model2')[term=="age",Estimate]

age_grid <- seq(0,21)


data.table(Age = age_grid,`E ~ age` = age_coef*age_grid,`E ~ age + D` = age_coef2*age_grid) %>% 
    pivot_longer(cols=c("E ~ age","E ~ age + D")) %>% 
    ggplot(aes(Age,value,color=name)) +
    geom_point() +
    geom_line() +
    guides(color=guide_legend(title="Model formula")) +
    xlab("Age") +
    ylab("Risk of Event")

```

## Polynomial Regression

```{r poly}
# http://www.science.smith.edu/~jcrouser/SDS293/labs/lab12-r.html
fit_plot_polynomial <- function(degree=3){
    fit = glm(Estate ~ poly(age, degree), data = report_dat)
    preds = predict(fit, newdata = list(age = age_grid), se = TRUE)
    pfit = exp(preds$fit) / (1+exp(preds$fit))
    se_bands_logit = cbind("upper" = preds$fit+2*preds$se.fit, 
                           "lower" = preds$fit-2*preds$se.fit)
    se_bands = exp(se_bands_logit) / (1+exp(se_bands_logit))
    
    E = report_dat[E=="Event"]
    
    notE = report_dat[E=="Other events"]
    
    coefs = sapply(coef(fit)[2:(degree+1)],function(x){round(x,2)})
    formula = 
        paste0("Event ~ ",paste0(sapply(1:degree,function(x){paste0(coefs[x]," * age^",x)}),collapse=" + "))
    
    ggplot() +
      geom_rug(data = notE, aes(x = jitter(age), y = max(pfit)), sides = "b", alpha = 0.1) +
      geom_rug(data = E, aes(x = jitter(age), y = min(pfit)), sides = "t", alpha = 0.1) +
      geom_line(aes(x = age_grid, y = pfit), color = "#0000FF") +
      geom_ribbon(aes(x = age_grid, 
                      ymin = se_bands[,"lower"], 
                      ymax = se_bands[,"upper"]), 
                  alpha = 0.3) +
      labs(title = paste0(formula,"\nAIC=",round(fit$aic,0)),
           x = "Age",
           y = "P(Event)")
}

fit_plot_polynomial(1)
fit_plot_polynomial(2)
fit_plot_polynomial(3)
fit_plot_polynomial(4)
fit_plot_polynomial(5)
fit_plot_polynomial(6)
fit_plot_polynomial(7)
```

## Regression splines

```{r splines_broken}

# Simon Wood's Intro to GAMs textbook chapter 4

tf <- function(x,xj,j){
    ## generate the jth tent function from set defined by knots xj
    dj <- xj*0;dj[j] <- 1
    ## linearly interpolates between xj and dj taking place at x
    approx(xj,dj,x)$y
}

tf.X <- function(x,xj){
    ## tent function basis matrix given data X
    ## and knot sequence xk
    nk <- length(xj); n <- length(x)
    X <- matrix(NA,n,nk)
    for(j in 1:nk) X[,j] <- tf(x,xj,j)
    X
}

knots = c(27/365,1,2,6,11,17,19)
Xp <- tf.X(age_grid,knots)

Xp_dt = data.table(Xp)
colnames(Xp_dt) <- knots %>% as.character()
Xp_dt$age <- age_grid
Xp_dt_melt <- 
    Xp_dt %>% 
    melt(id.vars="age",
         variable.name="knot")

Xp_dt_melt %>% 
    ggplot(aes(age,value,color=knot)) +
    geom_point() +
    geom_line() +
    facet_grid(knot~.)

knots = 1:length(stages)
Xp <- tf.X(knots,knots)

Xp_dt = data.table(Xp)
colnames(Xp_dt) <- knots %>% as.character()
Xp_dt$nichd <- knots
Xp_dt_melt <- 
    Xp_dt %>% 
    melt(id.vars="nichd",
         variable.name="knot")

Xp_dt_melt %>% 
    ggplot(aes(nichd,value,color=knot)) +
    geom_point() +
    geom_line() +
    facet_grid(knot~.) +
  xlab("NICHD developmental stage\n(term_neonatal=1 to late_adolescence=7")
```


```{r splines}
# https://stats.stackexchange.com/questions/29345/visualizing-a-spline-basis/29346
knots = c(27/365,1,2,6,11,17,19)
x <- report_dat$age
spl <- ns(x,knots=knots)

#sapply(1:ncol(spl),function(i){plot(x,spl[,i],ylab=paste0("Basis ",i),xlab="Age")})

autoplot(spl)

knots=c(1,2,3,4,5,6)
x <- report_dat$nichd %>% as.integer()
spl <- ns(x,knots=knots)

autoplot(spl) +
  xlab("NICHD developmental stage\n(term_neonatal=1 to late_adolescence=7")
```

## Generalized additive models

<!-- https://rdrr.io/cran/mgcv/man/smooth.terms.html --->

```{r gam_formulas}
gam_datas <- NULL

make_gam_data <- function(mod){
  
  data.table(
    term = names(coef(mod)), 
    coef = coef(mod), 
    se = sqrt(diag(vcov(mod))), 
    aic = mod$aic,
    formula = Reduce(paste, deparse(mod$formula))
    )
  
}

mod = bamV(E ~ D,
           data=report_dat,
           family=binomial(link="logit"),
           method = "fREML",
           aGam=list("discrete"=T))

gam_datas <- bind_rows(gam_datas,make_gam_data(mod))

summary(mod)

print(plot(mod,allTerms = T),pages=1)

mod = bamV(E ~ age,
           data=report_dat,
           family=binomial(link="logit"),
           method = "fREML",
           aGam=list("discrete"=T))

gam_datas <- bind_rows(gam_datas,make_gam_data(mod))

summary(mod)

print(plot(mod,allTerms = T),pages=1)

mod = bamV(E ~ s(age,
                 bs="cs"),
           data=report_dat,
           family=binomial(link="logit"),
           method = "fREML",
           aGam=list("discrete"=T))

gam_datas <- bind_rows(gam_datas,make_gam_data(mod))

summary(mod)

print(plot(mod,allTerms = T),pages=1)

mod = bamV(E ~ s(age,
                 bs="cs",
                 k=7),
           data=report_dat,
           family=binomial(link="logit"),
           method = "fREML",
           aGam=list("discrete"=T,"knots"=list(x=stage_knots_in_age)))

gam_datas <- bind_rows(gam_datas,make_gam_data(mod))

summary(mod)

print(plot(mod,allTerms = T),pages=1)

mod = bamV(E ~ s(age,
                 by=D,
                 bs="cs",
                 k=7),
           data=report_dat,
           family=binomial(link="logit"),
           method = "fREML",
           aGam=list("discrete"=T,"knots"=list(x=stage_knots_in_age)))

gam_datas <- bind_rows(gam_datas,make_gam_data(mod))

summary(mod)

print(plot(mod,allTerms = T),pages=1)

mod = bamV(E ~ nichd,
           data=report_dat,
           family=binomial(link="logit"),
           method = "fREML",
           aGam=list("discrete"=T))

gam_datas <- bind_rows(gam_datas,make_gam_data(mod))

summary(mod)

print(plot(mod,allTerms = T),pages=1)

mod = bamV(E ~ s(nichd_int,
                 bs="cs",
                 k=7),
           data=report_dat,
           family=binomial(link="logit"),
           method = "fREML",
           aGam=list("discrete"=T,"knots"=list(x=1:length(stages))))

gam_datas <- bind_rows(gam_datas,make_gam_data(mod))

summary(mod)

print(plot(mod,allTerms = T),pages=1) 

mod = bamV(E ~ s(nichd_int,
                 by=D,
                 bs="cs",
                 k=7),
           data=report_dat,
           family=binomial(link="logit"),
           method = "fREML",
           aGam=list("discrete"=T,"knots"=list(x=1:length(stages))))

gam_datas <- bind_rows(gam_datas,make_gam_data(mod))

summary(mod)

print(plot(mod,allTerms = T),pages=1)


```

```{r gam_metrics, fig.width=18,fig.height=10}

gam_datas %>% 
  .[order(term)] %>% 
  ggplot(aes(coef,term,color=formula)) + 
  geom_point(size=4) + 
  theme(legend.position = "none") + 
  facet_wrap(formula~.,scales="free")

gam_datas %>% 
  ggplot(aes(aic,formula)) + 
  geom_point(size=4)

```
